---
title: 'Reproducible Research: Peer Assessment 1'
author: "Belal Gooda"
output: html_document
---

## Loading Required Dependencies
```{r LoadingDependencies,message=FALSE,warning=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
```

## Loading and preprocessing the data

Unzipping the data file, and reading the data.

```{r DataLoading, cache=TRUE}
if(!file.exists('activity.csv')){
    unzip('activity.zip')
}
Activity <- read.csv("activity.csv", colClasses = c("numeric", "character", 
                                                    "numeric"))
```

Using dplyr and Lubridate packages to transform date column into date format.
```{r ConvertDate,cache=TRUE}
Activity <- Activity %>% rowwise() %>% mutate(date = ymd(date))
```


## What is mean total number of steps taken per day?
Calculating the total steps per day
```{r StepsPerDay,cache=TRUE, warning=FALSE}
StepsPerDay <- Activity %>% group_by(date) %>% 
        summarise(TotalSteps = sum(steps, na.rm = TRUE))
head(StepsPerDay)
```

Plotting a histogram of steps per day
```{r StepsPerDayHistogram, message=FALSE}
a <- ggplot(StepsPerDay, aes(TotalSteps))
a + geom_histogram(fill = "steelblue") +
        ggtitle("Total Steps Per Day") +
        xlab("Total Steps") + 
        ylab("Count") +
        theme_bw()
```

Calculating the mean and the median of total steps per day
```{r MeanMedian}
mean(StepsPerDay$TotalSteps)
median(StepsPerDay$TotalSteps)
```

From the above calculation, the mean is: 9354.23.  
While the median is: 10395.  


## What is the average daily activity pattern?

Calculating average steps per 5-minute interval across all days.
```{r AverageSteps,cache=TRUE, warning=FALSE}
AverageStepsPerInterval <- Activity %>% group_by(interval) %>% 
        summarise(AverageSteps = mean(steps, na.rm = TRUE))
head(AverageStepsPerInterval)
```

Creating a time series plot.
```{r AverageStepsTimeSeries, message=FALSE}
a <- ggplot(AverageStepsPerInterval, aes(x=interval,y=AverageSteps))
a + geom_line(color = "steelblue", size = 1) + 
        ggtitle("Daily Activity Pattern") + 
        xlab("5-minute Interval") + 
        ylab("Average Number of Steps") +
        theme_bw()
```

Finding The 5-minute interval with the maximum average number of steps.
```{r IntervalMaxAverage}
# Finding the index of the interval
IntervalIndex <- which.max(AverageStepsPerInterval$AverageSteps)
# Finding the value of the interval
IntervalValue <- AverageStepsPerInterval$interval[IntervalIndex]
```

We find that the 5-minute interval with the maximum average number of steps is:
`r IntervalValue`

## Imputing missing values

Calculating the number of missing values in the dataset
```{r NumberOfNAs}
SumNa <- sum(is.na(Activity))
```
The number of rows with missing values is: `r SumNa`

The strategy is to impute the steps missing values with the average steps of the correspnding 5-minute interval.

The following code, uses the dplyr package, and makes use of the previously computed "Average Steps Per Interval",
to create an imputed dataset.
```{r ImputingData, cache = TRUE}
ImputedActivity <- Activity %>% 
        mutate(steps = ifelse(
                is.na(steps),
                unlist(AverageStepsPerInterval[which(AverageStepsPerInterval$interval == interval),2]),
                steps))
```
As we see there was a bunch of missing values at the beginning of our activity dataset:
```{r}
head(Activity)
```
Which are not imputed with the correct averages:
```{r}
# Compare the values of the averages of the 5-minute intervals
head(AverageStepsPerInterval)
# To the value of the same 5-minute intervals in the imputed dataset
head(ImputedActivity)
```
Now we'll re-create the total steps per day histogram.
```{r StepsPerDayHistogramImputed, cache= TRUE, message=FALSE, warning=FALSE}
ImputedStepsPerDay <- ImputedActivity %>% group_by(date) %>% 
        summarise(TotalSteps = sum(steps, na.rm = TRUE))
a <- ggplot(ImputedStepsPerDay, aes(TotalSteps))
a + geom_histogram(fill = "steelblue") +
        ggtitle("Total Steps Per Day") +
        xlab("Total Steps") + 
        ylab("Count") +
        theme_bw()
```

The mean and median of the imputed total steps is:
```{r MeanMedianImputed}
mean(ImputedStepsPerDay$TotalSteps)
median(ImputedStepsPerDay$TotalSteps)
```

We notice that the mean and median are now the same after imputing.
The median did not change, and there is a little increase to the mean.

## Are there differences in activity patterns between weekdays and weekends?
For this part of the analysis, we will use the imputed dataset.  
We will first add a column for the week day names (DayName) corresponding to the dates. 
Then we will create a factor variable column (DayType), identifying weekdays and weekends.  
```{r DayNames, cache = TRUE}
ImputedActivity <- ImputedActivity %>%
        mutate(DayName = weekdays(date)) %>%
        mutate(DayType = ifelse(
                DayName %in% c('Saturday','Sunday'),
                'Weekend',
                'Weekday'
        ))
ImputedActivity$DayType <- factor(ImputedActivity$DayType)
# Now we can see the final dataset.
head(ImputedActivity)
# And we can see how many days are weekdays or weekends.
table(ImputedActivity$DayType)
```

Calculating the average steps per interval per weekday or weekend.  
```{r}
AverageStepsPerIntervalPerDayType <- ImputedActivity %>% group_by(interval,DayType) %>% 
        summarise(AverageSteps = mean(steps, na.rm = TRUE))
head(AverageStepsPerIntervalPerDayType)
```

Creating a time series plot of the average steps per interval for both types of days.
```{r}
a <- ggplot(AverageStepsPerIntervalPerDayType, aes(interval, AverageSteps))
a + geom_line(color = "steelblue", size = 1) + 
        facet_grid(DayType ~ .) + 
        ggtitle("Daily Activity Pattern") +
        xlab("5-minute Interval") + 
        ylab("Average Number of Steps") +
        theme_bw()
```

We notice from the above plot, that steps are much less in the start of the weekend day, 
compared to the start of a weekday.